---
title: "Building text prediction model"
subtitle: "Week 2 - Task 3"
author: "Sam Edwardes"
date: "2019-07-25"
output: html_notebook
---

```{r setup, warning=FALSE, message=FALSE, echo=FALSE}
# kintr settings
knitr::opts_knit$set(root.dir = "~/Documents/GitHub/predictive-text-model-swift-key")
```
## Background

The purpose of this document is to walk the reader through building of my text prediction model. The models works by:

- A "corpus" is created by combing twitter, news, and blog tests from the internet (data can be [downloaded here](https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip))
- A random sample of the corpus is taken to server as training data for the model. Additionally the entire data set is very large (over 1 GB), so using the entire dataset to build the model will result in a very slow application
- An n-gram data feature matrix is created for one to four gram sequences
- When text is entered, the model  attempts to find the longest matching n-gram
- The last word of the longest matching n-gram with the highest probabily based on the training data is returned as the prediction

## Environment

The following packages are required to build the model.

```{r warning=FALSE, message=FALSE}
# libraries
library(quanteda)
library(readtext)
library(tidyverse)
```

## Load the data

Read the text data from local disc. Combine all the files to create a single corpus document.

```{r}
# read all data
news_big <- readtext("data/raw/en_US.news.txt")
blogs_big <- readtext("data/raw/en_US.blogs.txt")
twitter_big <- readtext("data/raw/en_US.twitter.txt")

# create corpus
news_corp <- corpus(news_big)
blogs_corp <- corpus(blogs_big)
twitter_corp <- corpus(twitter_big)

# segment the corpus to lines
news_corp <- corpus_segment(news_corp, pattern = "\n")
blogs_corp <- corpus_segment(blogs_corp, pattern = "\n")
twitter_corp <- corpus_segment(twitter_corp, pattern = "\n")

# add all corpus together
all_corp <- news_corp + blogs_corp + twitter_corp

rm(news_corp, blogs_corp, twitter_corp,
   news_big, blogs_big, twitter_big)
```


## Clean the corpus

Remove questionable characters from the corpus that could impact the results.

```{r}
clean_corpus <- function(corp){
    
    # remove anything that is not alpha or number
    texts(corp) <- stringr::str_replace_all(texts(corp),"[^a-zA-Z\\s]", "")
    return(corp)
    
}
```


Select training and testing data.

```{r}
pop_size <- length(all_corp$documents$texts)

# define training data
set.seed(2019-07-24)
train_corp <- corpus_sample(all_corp, size = pop_size * 0.05)
train_corp <- clean_corpus(train_corp)

# define testing data
set.seed(2019-07-25)
test_corp <- corpus_sample(all_corp, size = pop_size * 0.001)
test_corp <- clean_corpus(test_corp)
```

## Process tokens

Create a function to apply to tokens to consistently clean them. This function will be used to clean the training data, as well as any text that a prediction will be made on.

```{r}

process_tokens <- function(toks){
    
    # create tokens
    toks <- tokens(toks, 
                   remove_numbers = TRUE, remove_punct = TRUE, remove_symbols = TRUE,
                   remove_separators = TRUE, remove_twitter = TRUE, remove_hyphens = TRUE,
                   remove_url = TRUE)
    
    # clean tokens
    toks <- tokens_tolower(toks)
    
    # get rid of stop words
    # toks <- tokens_select(toks, stopwords("english"), selection = "remove")
    
    # after some analysis, it was determined that stopwords should not be removed. 
    # It was helpful to remoe stopwords when you wanted to predict the answer to a question 
    # (e.g. who is the president of the united states?). 
    # But, it was not helpful for predicting what would by typed next
    
    # get rid of tokens that do not mean anything
    toks <- tokens_replace(toks, pattern = "â", replacement = "and")
    toks <- tokens_select(toks, pattern = "s", selection = "remove")
    
    # replace bad characters
    patterns <- c("â")
    toks <- as.tokens(lapply(toks, gsub, pattern = patterns, replacement = ""))
    
    # word stem
    toks <- tokens_wordstem(toks)
    
    return(toks)
}
```

Apply the tokens processing to our training data.

```{r}
# create the tokens
train.tokens <- tokens(train_corp)
train.tokens <- process_tokens(train.tokens)
```


## Build n-grams

Build ngrams ranging in size from 2 to 4.

```{r}
train.ngram1 <-tokens_ngrams(train.tokens, n = 1)
train.ngram2 <-tokens_ngrams(train.tokens, n = 2)
train.ngram3 <-tokens_ngrams(train.tokens, n = 3)
train.ngram4 <-tokens_ngrams(train.tokens, n = 4)
```


Turn into a data feature matrix (dfm), calculate frequencies, then turn into a data frame.

```{r}
# create a document feature matrix (dfm) for up to 4 ngrams
train.dfm1 <- textstat_frequency(dfm(train.ngram1))
train.dfm1$ngrams <- 1

train.dfm2 <- textstat_frequency(dfm(train.ngram2))
train.dfm2$ngrams <- 2

train.dfm3 <- textstat_frequency(dfm(train.ngram3))
train.dfm3$ngrams <- 3

train.dfm4 <- textstat_frequency(dfm(train.ngram4))
train.dfm4$ngrams <- 4

# combine all the dfms together
train.dfm <- rbind(train.dfm1, train.dfm2, train.dfm3, train.dfm4)

# Only keep n fetures with at least 2 observations
train.dfm <- train.dfm %>% filter(frequency >= 2)

# Check results
head(train.dfm, 20)

rm(train.dfm1, train.dfm2, train.dfm3, train.dfm4,
   train.ngram1, train.ngram2, train.ngram3, train.ngram4)
```

## The model

Two functions were created to execute the model:

**predict_next_word()**

- input: takes any string as an input and predicts what word will come next.
- returns: a list containing...
-- "input_text": the raw text entered into the search.
-- "input_tokens": the tokens that were used to perform the prediction.
-- "potential_matches": a dataframe containing the top 10 matches and probabilities.

**get_word**

- input: takes any string as an input and predicts what word will come next.
- returns: only the top predicted word.

**get_word3**

- same as get_word, but returns the top 3 results

```{r}
predict_next_word <- function(input_text){
    
    # process the input text
    input_toks <- tokens(input_text)
    input_toks <- process_tokens(input_toks)
    
    # determine the range of ngrams to search
    num_toks <- length(input_toks[[1]])
    ngram_range <- min(3, num_toks):1
    
    # loop through each range of ngrams to find matches
    potential_matches <- data.frame()
    for (num in ngram_range){
        
        # create the ngrams
        search_ngrams <- tokens_ngrams(input_toks, num)
        # create the search phrase
        search_term <- paste0("^", tail(search_ngrams[[1]], 1), "_")
        # identify the in scope search area
        search.dfm <- train.dfm %>% filter (ngrams >= num + 1)
        # search the dataframe
        temp <- search.dfm[grepl(pattern = search_term, x = search.dfm$feature), ]
        if(nrow(temp) == 0) next
        temp$ngram.size <- num
        # calculate frequncy and score
        total_ngrams <- sum(temp$frequency)
        temp <- temp %>%
            mutate(frequency.percent = frequency / total_ngrams) %>%
            mutate(score = num + frequency.percent)
        # join the results together
        potential_matches <- union_all(temp, potential_matches)
    }
    
    # check to see if there are matches
    if(nrow(potential_matches) == 0) {
        return(list("input_text" = input_text,
                    "input_toks" = input_toks,
                    "potential_matches" = NULL
    ))
    }
    
    # arrange the potential matches to find the most likely
    potential_matches <- potential_matches %>%
        arrange(desc(score), desc(frequency.percent)) %>% 
        head(10)
    
    # get the best predicted next word
    potential_matches$split_location <- stringi::stri_locate_last(str = potential_matches$feature, regex = "_")[,2]
    potential_matches <- potential_matches %>%
        mutate(prediction = substr(feature, split_location + 1, nchar(feature))) %>%
        select(feature, prediction, score, frequency.percent, ngram.size, ngrams, frequency)
    
    
    # return results
    return(list("input_text" = input_text,
                "input_toks" = input_toks,
                "potential_matches" = potential_matches
                ))
}

# Only return the top word
get_word <- function(x){
    y <- predict_next_word(x)
    if (is.null(y$potential_matches)){
        return("New word, no prediction")
    }
    return(y$potential_matches[1,2])
}

# Only return the top 3 words
get_word3 <- function(x){
    y <- predict_next_word(x)
    if (is.null(y$potential_matches)){
        return("New word, no prediction")
    }
    return(y$potential_matches[1:3,2])
}


# TESTING THE FUNCTIONS
test_text <- "Holy cow this prediction is going to be really"
z <- predict_next_word(test_text); z
get_word(test_text)
get_word3(test_text)
```

## Model accuracy

Build the test data.

```{r}
# tokenize test data
test.tokens <- tokens(test_corp)

# create the test data
test.text <- c()
test.correct.prediction <- c()
for (test_id in 1:length(test.tokens)){
    
    # figure out how many words in text
    num_words <- length(test.tokens[[test_id]]) - 1 # minus 1 so that there is always at least one word to predict
    if(num_words <= 1) next
    cutoff <- as.integer(runif(1, 1, num_words))
    
    temp.search <- paste(test.tokens[[test_id]][1:cutoff], collapse = " ")
    temp.prediction <- test.tokens[[test_id]][cutoff+1]
    
    # print(temp.search); print(temp.prediction)
    
    test.text <- append(test.text, temp.search)
    test.correct.prediction <- append(test.correct.prediction, temp.prediction)
}
```

Run the model on the test data.

```{r}
# Select how many tests you want to run
num_tests <- 500

# create a dataframe to hold the results
test_result_df <- data.frame(test.text[1:num_tests], test.correct.prediction[1:num_tests]);
names(test_result_df) <- c("test.text", "test.correct.prediciton")
rownames(test_result_df) <- 1:num_tests

# run the predction
prediction.result <- sapply(test.text[1:num_tests], get_word)
test_result_df <- cbind(test_result_df, prediction.result)

prediction.result.3 <- sapply(test.text[1:num_tests], get_word3)

prediction.result1 <- prediction.result.3[1:num_tests, 1]
prediction.result2 <- prediction.result.3[1:num_tests, 2]
prediction.result3 <- prediction.result.3[1:num_tests, 3]
test_result_df <- cbind(test_result_df, prediction.result1, prediction.result2, prediction.result3)

# clean the prediction results
test_result_df <- test_result_df %>%
    mutate(prediction.result = as.character(prediction.result),
           test.correct.prediciton = as.character(test.correct.prediciton),
           correct = prediction.result == test.correct.prediciton) %>%
   # mutate(correct = case_when(prediction.result == test.correct.prediciton ~ TRUE, TRUE ~ FALSE)) %>%
    select(prediction.result, test.correct.prediciton, correct, prediction.result1, prediction.result2, prediction.result3, test.text)



# summarise the results
num_correct <- sum(test_result_df$correct)
num_obs <- nrow(test_result_df)

print(paste0("Correct predictions: ", num_correct))
print(paste0("Total predictions: ", num_obs))
print(paste0("Accuracy rate: ", num_correct/num_obs))

test_result_df
```



```{r}
print("The end")
```

